APIs for hyperparameter optimization in scikit-learn

### 2013-05-12 paper.tex
focus on hyperparameter optimization

\documentclass[a4paper]{article}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage[round]{natbib}

% working title, please change
\title{Python as a machine learning language:
       experiences from the scikit-learn project}

@@

This paper discusses design choices made
in both the programmers' interface (API) and the implementation
of scikit-learn.
It also highlights obstacles faced by users and implementers
of machine learning algorithms in the Python language.
\end{abstract}

\section{Project vision}

\subsection{General principles}

@@

The main goal of the scikit-learn project
is to provide high-quality implementations
of ``classic'' machine learning algorithms
with a consistent API.

@@

@@

The bulk of the scikit-learn application programming interface
consists of so-called \textit{estimator} objects.
All learning algorithms for classification, regression and clustering
are offered as such objects,
but so are feature extraction, feature selection and dimensionality reduction.

@@

The algorithmic workhorse of most estimators, and one of the few methods
that all estimators have in common, is called \texttt{fit}.
Invoking this method causes the estimator to learn from a training set,
supplied as one or more arguments.
Its task, from an API point of view,
is to determine (via a learning algorithm) model-specific parameters
from a training set and set these as attributes on the estimator object,
which can then be used to transform data or make predictions.
As in other object-oriented machine learning libraries
(e.g.\ Weka's Java API)
no distinction is made in the type hierarchy
between what is variously called an estimator or a \textit{learner}
in the machine learning literature,
and the \textit{model} that results from applying an estimator to data.
When the \texttt{fit} method has been called,
an estimator object serves dual purpose as a model
(and is referred to as such in the documentation).

An example to clarify the partial application style
is the following code snippet.
If $X$ is a dataset in the format described below and $k$ is some integer,
then the Python code to train a $k$-means clustering model on it
can be written as
% XXX we could use Pygments for syntax highlighting
\begin{verbatim}
km = KMeans(n_clusters=k).fit(X)
\end{verbatim}
This snippet relies on the fact that the \texttt{fit} method
always returns the object it was called on (its ``\texttt{self}''),
making a method chaining style possible.

@@

\section{Pipelines and model selection}

A distinguishing feature of scikit-learn estimators
is that they can be composed into \texttt{Pipeline} objects.
Not only do such objects combine typical machine learning workflows
(extracting features, centering/scaling/normalization,
feature selection, making predictions)
in a single object that is itself an estimator,
and whose methods cause data to flow through the pipeline.
More importantly, though, they can be used for
systematic, automated model selection.
This is best explained by example.

>>>>

\documentclass[a4paper]{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage[round]{natbib}

% working title, please change
\title{APIs for hyperparameter optimization in scikit-learn}

@@

This paper discusses design choices made
in both the programmers' interface (API) and the implementation
of scikit-learn,
with a focus on the problem of hyperparameter optimization.
%It also highlights obstacles faced by users and implementers
%of machine learning algorithms in the Python language.
\end{abstract}

\section{Introduction}

\subsection{Project vision}

@@

The main goal of the scikit-learn project
is to provide high-quality implementations
of ``classic'' machine learning algorithms
with a consistent API.
This paper explores the role that such an API plays
for handling the problem of hyperparameter optimization,
or \textit{model selection} as it is called in the statistics literature.

The rest of this section explains some fundamental design choices
made throughout scikit-learn.
% TODO and the rest of the paper...

@@

\section{The hyperparameter optimization problem}

The focus of many machine learning toolkits is the \textit{learning} task,
which, at least in the supervised case,
can be formalized as the problem of finding a function
$\mathsf{model} : \mathcal{X} \to \mathcal{Y}$,
where $\mathcal{X}$ is some set of inputs
and $\mathcal{Y}$ some set of targets
($\mathbb{R}^k$ in regression, $\{1,\dots,K\}$ in classification).
An algorithm that finds such a function is called a \textit{learner}
% TODO: cite Mitchell 1998
or alternatively an \textit{estimator}
(in the scikit-learn documentation, and therefore in this paper).
Using function currying, its function signature is

\[
  L : \mathcal{H} \to \mathcal{X} \to \mathcal{Y}
\]

where $\mathcal{H}$ is the set of all possible \textit{hyperparameter} settings
for the estimator in question.\footnote{
  We disregard the fact that different estimators
  have widely varying sets of hyperparameters for notational simplicity.}
Hyperparameters include, for example,
regularization strength of linear models or SVMs,
choice of loss function for generic linear model learners,
or the number of features to keep in a feature selection method.
They are often treated as user-settable parameters
and not part of the learning problem proper,
either because they would affect the learning problem
in a way that would make it intractable \citep{bergstra2012}
or because their effects cannot properly be measured
in terms of a loss function on a training set.

However, in recent years, algorithms have been published
that systematically explore spaces of possible hyperparameter settings.
The problem of hyperparameter tuning
is crucial to achieving good generalization performance in machine learning.
The present paper does not go into hyperparameter optimization algorithms
in depth, focusing instead on API support for using them.

@@

The bulk of the scikit-learn application programming interface
consists of estimator classes.
All learning algorithms for classification, regression and clustering
are offered as such classes,
but so are feature extraction, feature selection and dimensionality reduction.
The three latter types of estimator are examples of a subtype
called \textit{transformers}:
in this case, both input and output are NumPy arrays representing samples,
but commonly with different numbers of columns.
Transformers that perform feature extraction on e.g.\ free text
are an exception in that their input is not usually a NumPy array.

@@

The algorithmic workhorse of most estimators, and one of the few methods
that all estimators have in common, is called \texttt{fit}.
Invoking this method causes the estimator to learn from a training set,
supplied as one or more arguments.
Its task, from an API point of view,
is to determine (via a learning algorithm) model-specific parameters
from a training set and set these as attributes on the estimator object,
which can then be used to transform data or make predictions.
As in other object-oriented machine learning libraries
(e.g.\ Weka's Java API)
no distinction is made in the type hierarchy
between an estimator and the model that results from applying it to data.
When the \texttt{fit} method has been called,
an estimator object serves dual purpose as a model.

As an example to clarify the partial application style,
consider a linear support vector machine (SVM).
% Rewrite example using LinearSVC
If $X$ is a dataset in the format described above and $k$ is some integer,
then the Python code to train a $k$-means clustering model on it
can be written as
% XXX we could use Pygments for syntax highlighting
\begin{verbatim}
km = KMeans(n_clusters=k).fit(X)
\end{verbatim}
This snippet relies on the fact that the \texttt{fit} method
always returns the object it was called on (its ``\texttt{self}''),
making a method chaining style possible.

@@

\section{Pipelines}

A feature of scikit-learn estimators
that makes effective hyperparameter optimization possible
is the fact that they can be composed into \texttt{Pipeline} objects.
Not only do such objects combine typical machine learning workflows
(extracting features, centering/scaling/normalization,
feature selection, making predictions)
in a single object that is itself an estimator,
and whose methods cause data to flow through the pipeline.
More importantly, though, they can be used for
systematic, automated model selection.
This is best explained by example.
